---
title: "DEA of TES"
output: html_notebook
---

Load packages

```{r}
library(tidyverse)
library(DESeq2)
```


Reading in metadata

```{r}
getwd() # is project folder

# the sample metadata for both male and female is in a .csv in the data folder

metadata <- read.csv("../data/metadata/mouse_TE_metadata.csv")

# inspect
metadata
dim(metadata)
str(metadata)
```


Read in count data

```{r}
# Point to folder with the count .tsv file for each sample
folder_path <- "../data/counts/"
files_path <- list.files(path = folder_path, pattern = "\\.tsv$", full.names = T) # makes list
files_list <- lapply(files_path, read_tsv)

# Make into dataframe
str(files_list)
count_data_allsamples <- files_list %>% purrr::reduce(full_join, by='transcript') # There are NAs which can be dealt with later
# make the Geneid the row id, turns into matrix
count_data_allsamples <- count_data_allsamples %>% remove_rownames %>% column_to_rownames(var="transcript")
head(count_data_allsamples)
```

Make metadata row names match and reorder samples for both so they match 

```{r}
metadata <- metadata %>% remove_rownames %>% column_to_rownames(var="sample") # make row names
all(colnames(count_data_allsamples) %in% rownames(metadata)) # Check all samples are there

# make rownames/colnames in same order
count_data_allsamples <- count_data_allsamples[, rownames(metadata)]
all(rownames(metadata) == colnames(count_data_allsamples)) # True

# Replace all NA values with 0 
count_data_allsamples[is.na(count_data_allsamples)] <- 0
```


Create DESeq object

```{r}
dds <- DESeqDataSetFromMatrix(countData = count_data_allsamples,
                              colData = metadata,
                              design = ~ sex + treatment) # Treatment is last as it's the main thing we want to look at
```

Filter low count genes, based on smallets group size per DESeq2 vignette

```{r}
#prefilter 
smallestGroupSize <- 5 # sex/treatment
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize 
dds <- dds[keep,] # 56965 genes remaining
```

Relevel to ensure comparisons are being made relative to sham level

```{r}
dds$treatment <- relevel(dds$treatment, ref = 'SHAM')
```


Normalise the count data:

```{r}
# normalise 
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized = T)
# write normalised counts out to file
#write.table(normalized_counts, file="../data/normalised_counts_allsamples.tsv", sep = '\t', quote=F, col.names = NA)
```


Differential expression analysis
```{r}
dds <- DESeq(dds)
res <- results(dds, alpha = 0.05)
summary(res) # this looks much better, but still have questions


# write out full results
#write.csv(as.data.frame(res), file="../data/DE_TEs.csv")

```


Visualising results

```{r}

# visualising results
DE_TEs <- data.frame(res)
head(DE_TEs)
dim(DE_TEs)

# write out significant results
DE_TEs <- data.frame(res)
head(DE_TEs)
signif_TEs <- DE_TEs %>%
  filter(padj < 0.05)
write.csv(signif_TEs, file="../results_data/signif_DE_TEs.csv")


# rlog transform for purposes of PCA transformation
rld <- rlog(dds)
TE_PCA <- plotPCA(rld, intgroup=c("treatment", "sex")) # separation of sex, some separation of treatment, more in females

# customise pca??

#pcaData <- plotPCA(rld, intgroup=c("condition", "type"), returnData=TRUE)
#percentVar <- round(100 * attr(pcaData, "percentVar"))
#ggplot(pcaData, aes(PC1, PC2, color=condition, shape=type)) +
  #geom_point(size=3) +
  #xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  #ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  #coord_fixed()

ggsave("../figures/TE_PCA.png", TE_PCA, dpi = 300)
```

## interaction, shrinkage??

Check dispersion

```{r}
# dispersion plot
plotDispEsts(dds)

```

Volcano plot

```{r}
library(ggplot2)

# Add column for labelling purposes with either up/downregulated or not significant
DE_TEs$signif <- ifelse(DE_TEs$padj < 0.05 & abs(DE_TEs$log2FoldChange) > 0.5, ifelse(DE_TEs$log2FoldChange > 0, "Up", "Down"), "NotSig") # made LFC threshold low to capture downreg genes

# Omit NAs in the padj column, they are low counts and outliers, but why so many?
DE_TEs %>% filter(if_any(everything(), is.na)) %>%
  nrow() # 5534 rows containing NAs
DE_TEs <- na.omit(DE_TEs)
                                     
DE_TEs %>%
  filter(padj < 0.05) # 15 signif genes with LFC over 0.5

# Make labels for the plot
library(ggrepel)
# create a datframe just holding the signif genes
signifgenes <- DE_TEs %>%
  filter(padj < 0.05)
signifgenes
# for labelling
DE_TEs$label = if_else(rownames(DE_TEs) %in% rownames(signifgenes), rownames(DE_TEs), "")

# Plot
volcano <- ggplot(DE_TEs, aes(x = log2FoldChange, y = -log10(padj), color = signif)) +
geom_point()+labs(x = "Log2 Fold Change", y = "-Log10 p-value",title = "Differentially expressed TEs") +theme(legend.position = "right")+ scale_colour_manual(name = "Expression", labels = c("Downregulated", "Not significant", "Upregulated"), values = c("Down" = "blue", "NotSig" = "darkgrey", "Up" = "red"))+geom_text_repel(aes(label = label), size = 2)

volcano
ggsave("../figures/TE_volcano_plot.png", volcano, dpi = 300)

```

Make heatmap

```{r}
library("pheatmap")

sampleDists <- dist(t(assay(rld)))

select <- order(rowMeans(counts(dds,normalized=TRUE)),
                decreasing=TRUE)[1:20]
df <- as.data.frame(colData(dds)[,c("treatment","sex")])
ntd <- normTransform(dds)
heatmap <- pheatmap(assay(ntd)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)

ggsave("../figures/TE_heatmap.png", heatmap, dpi = 300)
```






